\documentclass[15pt,usletter,fancy]{src/phdthesis}
\usepackage{tikz}
\usepackage{titlesec}
\usepackage{setspace} % spaces between lines
\usepackage{parskip} % spaces between paragraphs
\usepackage{xcolor}
\usepackage{tcolorbox}


%\title{}
%\subtitle{}
%
%\author{Zhenhua Zhang}
%\institute{University Medical Center Groningen}
%\date{\today}
%\version{4.1}
%\bioinfo{Bio}{Information}

%\extrainfo{}

%\logo{logo-blue.png}
%\cover{cover.jpg}

% Set the line spacing to 1
% \onehalfspacing


\begin{document}

\chapter*{Prepositions}
\addcontentsline{toc}{chapter}{Prepositions}

\begin{enumerate}
  \item Host genetic and epigenetic regulations can be essential factors that determine the outcomes of viral infection. (\textit{this thesis})
  \item Genetic variants can modulate the outcomes of viral infection by affecting the epigenetic characteristics, e.g., chromatin accessibility. (\testit{this thesis})
  \item The integration of single-cell multi-omics and genetics provides novel insights into the regulatory cascades under acute viral infections. (\textit{this thesis})
  \item The activities of latent HIV-1 reservoirs are under host genetic regulations, which provides new opportunities in developing therapy strategies for HIV-1 infection. (\text{this thesis})
  \item Cell-type specific quantitative trait loci dissect heterogeneity of genetic regulation on abundance of cell surface proteins. (\textit{this thesis})
  \item Applications of machine (deep) learning algorithms in life sciences improve our understanding of the relationship between gene expressions and genomic architectures. (\textit{this thesis})
  \item The combination of comprehensive database / biobanks and advanced data-mining methods enables a bioinformatician to stand firmly and walk freely.
  \item As we start by asking biological questions, then explore them with multi-omics techniques and well-tuned algorithms, we either end with new insights or enter the endless exploratory odyssey.
  \item The best morality is like water. The water is good at benefiting all creatures but never establishes superiority over them. (\textit{Tao Te Ching}, translated by Zhenhua Zhang)
\end{enumerate}

\end{document}
