\documentclass[12pt]{letter}
\usepackage{geometry}

\geometry{margin=1in}

\begin{document}

Morris A Swertz, , +31652606501, m.a.swertz@rug.nl

\vspace{1cm}

\today

xxx xx \\
HZI

\vspace{1.5cm}

To Whom It May Concern:

As Zhenhua Zhang's primary supervisor, it is my pleasure to recommend him for the post-doc researcher under ``Friends of HZI'' fellowships.

I am Morris Swertz, a full professor in bioinformatics and biomedicine from Department of Genetics at University Medical Center Groningen and University Gronigen (NL).
As one one of my Ph.D. students, Zhenhua started to pursue his Ph.D. degree since April 2018 in my group and expects to finalize his thesis and receive his promotion in the beginning of 2022.

During the past almost four years, Zhenhua has been working on several projects to interpret the genetic mechanisms of human phenotypes.
These works yield decent and high quality publications in peer-reviewed scientific journals.
Of note, he developed many data analysis skills such as multi-omics data analysis, quantitative traits loci analysis, and machine-/deep-learning techniques.
With these skills, I believe that he is a qualified young scientist to be in charge of bioinformatic projects involving multi-omics data.

Zhenhua has positively developed himself since he started in our team in more than three years ago.
In the first several months he quickly accumulated the essential knowledge of our research domain, i.e., bioinformatics for human genetics especially and RNA sequencing.
Then he demonstrated that he can work disciplined and independently on difficult research assignments, while systematically reporting on the progress made, challenges encountered, and solutions proposed.
Also, I was happy to observe, that Zhenhua has been developing new research proposals on his own, and seeking collaboration with other researchers.
I am confident that Zhenhua is an independent researcher with strong ability to collaborate, as in our field multidisciplinary collaboration is key to success.

If you need more information or have any question about Zhenhua, please do not hesitate to contact me by email at \textit{m.a.swertz@rug.nl} or by phone +31652606501.
I would be happy to elaborate on my working time with him.

Sincerely,

Morris Swertz \\
Department of Genetics \\
University Medical Center Groningen

\end{document}
